Mutiple level calculation 2

### 2023-04-15 posts/rtables/multiple_levels_calculation_2.qmd
Create multiple_levels_calculation_2.qmd

---
title: 'Mutiple level calculation 1'
date: '2023-04-07'
author: Qing Zou
categories: ['R', 'rtables']
description: 'Explore how to handle the mutiple calculation by rtables'
execute:
  message: false
  warning: false
editor_options:
  chunk_output_type: console
---

@@

## Multiple lelvels calculation

`Multipile levels` calculation for categorical variable is the most common cases in Table, e.g. `PCSA`, `Exposure tables`.

In this section, we will explore how to produce these output in `rtables`.

Below is the test data:

```{r cal2}
#| echo: true
#| eval: true
df <- data.frame(SUBJID = c(rep('S1', 5), rep("S2", 4), rep("S3", 2)),
                 ASEQ = c(1:5, 1:4, 1:2),
                 GROUP = factor(c(rep('A', 5), rep("B", 4), rep('A', 2)), levels = c("A", "B")),
                 PARAMCD = c('TEST_1', 'TEST_1', 'TEST_1', 'TEST_1', 'TEST_2', 'TEST_1', 'TEST_1', 'TEST_2', 'TEST_2', 'TEST_1', 'TEST_2'),
                 CRIT = c('LEVEL_2_1','LEVEL_2_1', 'LEVEL_2_2', NA_character_, 'LEVEL_2_a', 'LEVEL_2_1', NA_character_, 'LEVEL_2_a', 'LEVEL_2_b', NA_character_, 'LEVEL_2_b'))
```
```{r cal2sd }
#| echo: false
#| eval: true
#| message: FALSE
knitr::kable(df, "html")%>%
  kableExtra::kable_styling(font_size = 16)
```

Now, we need to take the `PARAMCD` as 1st level, and the `CRIT` as 2nd level to calculate the frequency of each `TEST`, it will be a table like PCSA.

# Method 1

the `rtables::analyze()` will firstly be used in this case.

To use this function, the data should be transformed from `long strcture(BDS)` to `wide`.

Steps:

1. take action on setting up for the `missing value` and the `factor`:

 - fill the missing value as `Missing_value_flag`, which will be subsequently handled in `rtables::analyze()`
 - use function `tidyr::pivot_wider` implement the transformation, the new columns are used as the analyzed variables `vars` of `rtables::analyze()`
 - `factor` the values by given level values, the levels is the format of output

```{r cal2df}
#| echo: true
#| eval: true

@@

df1 <- df1 %>%
  mutate(across(paste0('TEST_', 1:2), ~factor(fmt[[cur_column()]][.x],
                                              levels = c(fmt[[cur_column()]])), .names = "{.col}_fmt"))
```
```{r cal2g }
#| echo: false
#| eval: true
#| message: FALSE
knitr::kable(df1, "html")%>%
  kableExtra::kable_styling(font_size = 16)
```

2. The dataframe `df1` is ready for output. now, the `layout` and `key values` of table will be built in the `rtables::analyze()`.

 - the header and overall column are generated by  `split_cols_by` and `add_overall_col`
 - in the `rtables::analyze()`:
   - `vars` and `var_labels` are the variable names and labels
   - in the `afun`, we create an analysis function, in which the arguments `df`, `.N_col`, `.var`, `.N_row` are created by rtables itself. the other aruments are customized will be passed by `extral_args`
     - the `df` is current available(`non NA`) data with splited group, that's why we firstly fill the missing data `Missing_value_flag`. By filling
       a non-NA value, we can handle the missing value as users expected.
     - you will see the customized arguments `missing_rm` and `missing_label`, if user hope hide or remove the missing values.
     - the `n_unique_keys` and `N_unique_keys` indicate that when counting the number of n/N, which records should be used.
     - `add_Number` means whether add an row `Number: N`
     - All the values of these customized arguments are passed by the `extra_args`, it's the greatest way to add the flexiabiliy for the tabulation.

```{r cal2as}
#| echo: true
#| eval: true

lyt <- basic_table() %>%
  split_cols_by("GROUP") %>%
  add_overall_col("All") %>%
  rtables::analyze(vars = c('TEST_1_fmt', 'TEST_2_fmt'),
                   var_labels = c('This is test 1', 'This is test 2'),
                   afun = function(df, .N_col, .var, .N_row,
                                  format = 'n(%)',
                                  decimal = 1,
                                  n_unique_keys = NULL,
                                  N_unique_keys = NULL,
                                  show_N_item = FALSE,
                                  add_Number = FALSE,
                                  missing_rm = TRUE,
                                  missing_label =NULL,
                                  ...) {

@@

                    df_n <- df
                    if(!is.null(n_unique_keys)){
                      df_n <- df %>% distinct(across(c(n_unique_keys, .var)), .keep_all = TRUE)
                    }

                    x <- df_n[[.var]]

                    if(!is.null(N_unique_keys)){
                      cus_N_col <- df %>% distinct(across(N_unique_keys)) %>% nrow()
                    }else{
                      cus_N_col <- .N_col
                    }

                    lst_body <- lapply(
                      as.list(table(x)),
                      function(xi) {
                        rcell(format_values_helper(xi, cus_N_col, type = format),
                              format = sprintf_format_helper(xi, cus_N_col, 1, type = format),
                              indent_mod = 1L
                        )
                      }
                    )

@@

                    if(add_Number){
                      lst_body <- append(setNames(list(rcell(cus_N_col,
                                                            format = "xx",
                                                            indent_mod = 1L)),
                                                  'Number'), lst_body)
                    }
                    in_rows(
                      .list = c(
                        lst_body,
                        # add empty row, in .list obj
                        " " = rcell("", format = "xx")
                      )
                    )
                  },
          extra_args = list(format = 'n/N(%)',
                            n_unique_keys = c("SUBJID"),
                            N_unique_keys = c("SUBJID"),
                            add_Number = FALSE,
                            missing_rm = FALSE,
                            missing_label = NULL)
          # ,
          # show_labels = "hidden"
  )
```
3. finally, fit the `layout` and `df1` into the `rtables::build_table`

```{r cal2ad}
#| echo: true
#| eval: true
col_counts_df <- df1 %>%
  distinct(SUBJID, GROUP)
col_counts <- table(col_counts_df$GROUP)
col_counts <- c(col_counts, sum(col_counts))

@@

```{r cap1}
#| echo: false
#| eval: true
my_caption <- ''
```

>>>>

---
title: 'Mutiple level calculation 2'
date: '2023-04-09'
author: Qing Zou
categories: ['R', 'rtables']
description: 'Wrap the analysis function'
execute:
  message: false
  warning: false
editor_options:
  chunk_output_type: console
---

@@

```{r cal2}
#| echo: false
#| eval: true
df <- data.frame(SUBJID = c(rep('S1', 5), rep("S2", 4), rep("S3", 2)),
                 ASEQ = c(1:5, 1:4, 1:2),
                 GROUP = factor(c(rep('A', 5), rep("B", 4), rep('A', 2)), levels = c("A", "B")),
                 PARAMCD = c('TEST_1', 'TEST_1', 'TEST_1', 'TEST_1', 'TEST_2', 'TEST_1', 'TEST_1', 'TEST_2', 'TEST_2', 'TEST_1', 'TEST_2'),
                 CRIT = c('LEVEL_2_1','LEVEL_2_1', 'LEVEL_2_2', NA_character_, 'LEVEL_2_a', 'LEVEL_2_1', NA_character_, 'LEVEL_2_a', 'LEVEL_2_b', NA_character_, 'LEVEL_2_b'))
```

```{r cal2df}
#| echo: false
#| eval: true

@@

df1 <- df1 %>%
  mutate(across(paste0('TEST_', 1:2), ~factor(fmt[[cur_column()]][.x], levels = c(fmt[[cur_column()]])), .names = "{.col}_fmt"))

col_counts_df <- df1 %>%
  distinct(SUBJID, GROUP)
col_counts <- table(col_counts_df$GROUP)
col_counts <- c(col_counts, sum(col_counts))
```

## Wrap the analysis function

The function is too long and it's not convient mixed with the layout of rtbales.
For better user experience, we can create a function `summarise_vars_categorical`

1. wrap the analysis function as `summarise_vars_categorical_afun`
2. wrap the `summarise_vars_categorical_afun` and the `extral_args` within the `summarise_vars_categorical`

Then user can passed all the analysis variables and customized argument by a single function `summarise_vars_categorical`.

```{r a23s}
#| echo: true
#| eval: true
#| message: FALSE

# wrap the analysis function
summarise_vars_categorical_afun <- function(df, .N_col, .var, .N_row,
                                            format = 'n(%)',
                                            decimal = 1,
                                            n_unique_keys = NULL,
                                            N_unique_keys = NULL,
                                            show_N_item = FALSE,
                                            add_Number = FALSE,
                                            missing_rm = TRUE,
                                            missing_label = NULL,
                                            indent_mod = 0L,
                                            ...) {

@@

  df_n <- df
  if(!is.null(n_unique_keys)){
    df_n <- df %>% distinct(across(c(n_unique_keys, .var)), .keep_all = TRUE)
  }
  x <- df_n[[.var]]
  if(!is.null(N_unique_keys)){
    cus_N_col <- df %>% distinct(across(N_unique_keys)) %>% nrow()
  }else{
    cus_N_col <- .N_col
  }

  lst_body <- lapply(
    as.list(table(x)),
    function(xi) {
      rcell(format_values_helper(xi, cus_N_col, type = format),
            format = sprintf_format_helper(xi, cus_N_col, 1, type = format),
            indent_mod = indent_mod
      )
    }
  )
  if(add_Number){
    lst_body <- append(setNames(list(rcell(cus_N_col,
                                           format = "xx",
                                           indent_mod = indent_mod)), 'Number'), lst_body)
  }

@@

  in_rows(
    .list = c(
      lst_body,
      # add empty row, in .list obj
      " " = rcell("", format = "xx")
    )
  )
}

# wrap the summarise_vars_categorical_afun and the extral_args
summarise_vars_categorical <- function(lyt,
                                       vars,
                                       var_labels,
                                       format = 'n(%)',
                                       decimal = 1,
                                       n_unique_keys = NULL,
                                       N_unique_keys = NULL,
                                       show_N_item = FALSE,
                                       add_Number = FALSE,
                                       missing_rm = TRUE,
                                       missing_label = NULL,
                                       show_labels='visible',
                                       indent_mod = 0L, ...){
  rtables::analyze(lyt, vars, var_labels,
                   afun = summarise_vars_categorical_afun,
                   extra_args = list(format = format,
                                     decimal = decimal,
                                     n_unique_keys = n_unique_keys,
                                     N_unique_keys = N_unique_keys,
                                     show_N_item = show_N_item,
                                     add_Number = add_Number,
                                     missing_rm = missing_rm,
                                     missing_label = missing_label),
                   show_labels = show_labels, indent_mod = indent_mod)
}

lyt <- basic_table() %>%
  split_cols_by("GROUP") %>%
  add_overall_col("All") %>%
  summarise_vars_categorical(vars = c('TEST_1_fmt', 'TEST_2_fmt'),
                             var_labels = c('This is test 1', 'This is test 2'),
                             format = 'n/N(%)',
                             n_unique_keys = c("SUBJID"),
                             N_unique_keys = c("SUBJID"),
                             add_Number = TRUE,
                             missing_rm = FALSE

  )

@@
